\def\pathToRoot{../../}\input{\pathToRoot headers/uebungHeader}

\begin{document}

% Use Basis x or Talk x, where x is the number of the session
\uebunghead{Talk 6}{Special Constructions in Categories}
\author{Maximilian Wuttke}

\hint{Read the slides for the definitions of the six constructions.}

\section{Products and Coproducts}

\begin{exercise}[Product is unique up-to isomorphism]
  Prove that for every category $\cat{A}$ and every objects $X, Y \in \cat{A}$
  all products of $X$ and $Y$ are isomorphic. Thus we can speak of \emph{the} product of $X$ and $Y$.
\end{exercise}

\begin{answer}
  \begin{proof}
    Let $X, Y \in \cat{A}$ and $P$ be a product of $X$ and $Y$ together with $p_1 \from P \to X, p_2 \from P \to Y$
    and $P'$ be a product of $X$ and $Y$ together with $p'_1 \from P' \to X, p'_2 \from P' \to Y$.

    Lets reconcider the definition of the \emph{product}:
    For every object $A \in \cat{A}$ and arrows $f_1 \from A \to X, f_2 \from A \to Y$ we get that there exists a unique map
    $\bar{f} \from A \to P$, such that
    \[ \xymatrix{
        &A \ar[ldd]_{f_1} \ar@{.>}[d]|{\bar{f}\vphantom{\bar{\bar{f}}}}
        \ar[rdd]^{f_2}&       \\
                &P \ar[ld]^{p_1} \ar[rd]_{p_2}                  &       \\
        X       &                                               &Y
    } \]
    commutes.
    Now we can instantiate $A$ with $P'$, $f_1$ with $p_1$, $f_2$ with $p_2$. By analogous instantiations we get the commuting diagram
    \[ \xymatrix{
      & P \ar@/_/[ldd]_{p_1} \ar@/^/[rdd]^{p_2} & \\
      & P \ar[ld]^{p_1} \ar@/_/@{.>}[dd]|{\bar{f}} \ar@{.>}[u]|{f} \ar[rd]_{p_2} & \\
      X & & Y \\
      & P' \ar[lu]_{p_1'} \ar@/_/@{.>}[uu]|{\tilde{f}} \ar@{.>}[d]|{f'} \ar[ru]^{p'_2} & \\
      & P' \ar@/^/[luu]^{p'_1} \ar@/_/[ruu]_{p'_2} &
    } \]
    for exactly one $f, f', \tilde{f}, \bar{f}$. With the inner triangles we get:
    \begin{alignat*}{2}
      p_1  \of \tilde{f} &= p'_1 &\qquad p_2  \of \tilde{f} &= p'_2 \\
      p'_1 \of \bar{f}   &= p_1  &       p'_2 \of \bar{f}   &= p_2
    \end{alignat*}
    By substitution we get
    \begin{alignat*}{2}
      p_1   \of (\tilde{f} \of \bar{f}) &= p_1  &\qquad p_2  \of (\tilde{f} \of \bar{f}) &= p_2 \\
      p'_1  \of (\bar{f} \of \tilde{f}) &= p'_1 &       p'_2 \of (\bar{f} \of \tilde{f}) &= p'_2
    \end{alignat*}
    With the outer triangles we get:
    \begin{alignat*}{5}
      p_1  &\;\of&&\; f  &= p_1  &\qquad p_2  &\;\of&\; f  &= p_2 \\
      p'_1 &\;\of&&\; f' &= p'_1 &\qquad p'_2 &\;\of&\; f' &= p'_2
    \end{alignat*}
    Note that $f$ and $f'$ are \emph{unique} and the first equations hold for $f = 1_P$ and $f = \tilde{f} \of \bar{f}$, thus $\tilde{f} \of \bar{f} = 1_P$.
    The second equations holds for $f' = 1_{P'}$ and for $f' = \bar{f} \of \tilde{f}$, thus $\bar{f} \of \tilde{f} = 1_{P'}$.
    Thus $P$ and $P'$ are isomorphic. \qedhere
  \end{proof}
\end{answer}

\hint{We will now assume that all the 6 constructions are unique up-to isomorphism.}

\begin{definition}{Arbitrary\ product}
  Let $\cat{A}$ be a category, $I$ a set, and $(X_i)_{i \in I}$ a family of objects of $\cat{A}$.
  A \demph{product} of $(X_i)_{i \in I}$ consists of an object $P$ and a family of maps
  $\Bigl(P \toby{p_i} X_i\Bigr)_{i \in I}$
  with the property that for all objects $A$ and families of maps
  $\Bigl(A \toby{f_i} X_i\Bigr)_{i \in I}$
  there exists a unique map $\bar{f}\from A \to P$ such that
  $\forall i \in I: p_i \of \bar{f} = f_i$.
\end{definition}

\begin{exercise}[Greatest lower bound]
  Define the \emph{greatest lower bound} for a family of objects in a poset and prove that it is the arbitrary product of the family when the poset is seen as a category.
\end{exercise}

\begin{answer}
  Let $(A, \le)$ be a poset concidered as a category and $I$ a set.
  A \demph{lower bound} of a family $(x_i)_{i\in I}$ is an element $a \in A$, s.\,t. $\forall i \in I: a \le x_i$.
  $z \in A$ is a \demph{greatest lower bound} of this family if $\forall a \in A: \text{$a$ is a lower bound} \Rightarrow a \le z$.

  Let $z$ be a greatest lower bound of the family $(x_i)_{i\in I}$. We show that $z$ is a product of $(x_i)_{i\in I}$.
  \begin{proof}
    We first have to define a family of maps $\Bigl(z \toby{p_i} x_i\Bigr)_{i \in I}$.
    This is trivial since we have $\forall i \in I: z \le x_i$.
    Now let $a \in A$ be an object and let $\Bigl(a \toby{f_i} x_i\Bigr)_{i \in I}$ be a family of maps from $a$.
    The existence of this maps means that $a$ is an upper bound of $(x_i)_{i\in I}$.
    Now by definition of least upper bound, $a \le z$, thus we have an arrow $\bar{f} \from a \to z$.
    All arrows are unique, thus $\bar{f}$ is the only map making the diagram commute. \qedhere
  \end{proof}
\end{answer}

\begin{definition}{Arbitrary\ coproduct}
  Let $\cat{A}$ be a category, $I$ a set, and $(X_i)_{i \in I}$ a family of objects of $\cat{A}$.
  A \demph{coproduct} of $(X_i)_{i \in I}$ consists of an object $S$ and a family of maps
  $\Bigl(X_i \toby{p_i} S\Bigr)_{i \in I}$
  with the property that for all objects $A$ and families of maps
  $\Bigl(X_i \toby{f_i} A\Bigr)_{i \in I}$
  there exists a unique map $\bar{f}\from S \to A$ such that
  $\forall i \in I: \bar{f} \of p_i = f_i$.
\end{definition}

\begin{exercise}[Disjoint union]
  \begin{itemize}
    \item[(a)]For an index set $I$ and a family of sets $(X_i)_{i \in I}$, define the arbitrary disjoint union $\coprod_{i \in I}{X_i}$.
    \item[(b)]Show that this arbitrary disjoint union is indeed the arbitrary coproduct in $\Set$.
  \end{itemize}
\end{exercise}

\begin{answer}
  First we define the arbitrary disjoint union:
  \begin{definition}{Arbitrary disjoint union}
    Let $I$ be an index set and let $(x_i)_{i \in I}$ be a family of sets.
    We define the arbitrary disjoint union of $(x_i)_{i\in I}$:
    $$\coprod_{i \in I} {x_i} := \bigcup_{i \in I} \set{ (x, i) \;\big|\; x \in x_i}$$
  \end{definition}
  Now let $I$ be an index set and let $(x_i)_{i\in I}$ be a family of sets. We show that $\coprod_{i \in I} {x_i}$ is the coproduct of $(x_i)_{i\in I}$.
  \begin{proof}
    We first define the projection maps: $p_i(a) := (a, i)$.
    Now let $A$ be a set and $\Bigl(x_i \toby{f_i} A\Bigr)_{i \in I}$ be a family of functions to $A$.
    We define $\bar{f} : \coprod_{i \in I} {x_i} \to A$ with $\bar{f}(a, i) := f_i(a)$.
    We have to show that $\bar{f}$ is the only function such that
    $\forall i \in I: \bar{f} \of p_i = f_i$.
    Let $i \in I$ and we check $a \in x_i$. $\bar{f}(p_i(a)) = \bar{f}(a,i) = f_i(a)$.
    Now let $\tilde{f} : \coprod_{i \in I} {x_i} \to A$ be another function such that
    $\forall i \in I: \tilde{f} \of p_i = f_i$.
    Let $(a, i) \in \coprod_{i \in I} {x_i}$. $\bar{f}(a,i)= f_i(a) = \tilde{f}(p_i(a)) = \tilde{f}(a,i)$.\qedhere
  \end{proof}
\end{answer}

\begin{exercise}[Direct sum of vector spaces]
  In the talk we have seen that the direct sum $V \oplus W$ of two vector spaces $V, W$ over a fixed field $K$ is the coproduct of $V$ and $W$.
  Show that it is also the product.
\end{exercise}

\begin{answer}
  \begin{proof}
    We first define the projection functions $p_1 \from V \oplus W \to V, p_2 \from V \oplus W \to W$:
    \begin{alignat*}{1}
      p_1(x,y) &:= x \\
      p_2(x,y) &:= y
    \end{alignat*}
    Clearly $p_1$ and $p_2$ are linear functions.
    Now let $A$ be a vector space and $f_1 \from A \to V, f_2 \from A \to W$ be linear functions.
    We define $\bar{f}(a) := (f_1(a), f_2(a))$. We clearly have
    \begin{alignat*}{2}
      p_1(\bar{f}(a)) &= p_1(f_1(a), f_2(a)) = f_1(a) \\
      p_2(\bar{f}(a)) &= p_2(f_1(a), f_2(a)) = f_2(a).
    \end{alignat*}
    Thus the diagram
    \[ \xymatrix{
        & A \ar[ldd]_{f_1} \ar@{.>}[d]|{\bar{f}\vphantom{\bar{\bar{f}}}} \ar[rdd]^{f_2} & \\
        & {V \oplus W} \ar[ld]^{p_1} \ar[rd]_{p_2} & \\
        V & & W
    } \]
    commutes. Now we have to show that $\bar{f}$ is the only map such that the above diagram commutes.
    Let $\tilde{f} \from A \to V \oplus W$ be a map such that $p_1 \of \tilde{f} = f_1 \land p_2 \of \tilde{f} = f_2$.
    \[
      \tilde{f}(a) = \bigl(p_1(\tilde{f}(a)), p_2(\tilde{f}(a))\bigr) = \bigl(f_1(a), f_2(A)\bigr) = \bar{f}(a)\qedhere
    \]
  \end{proof}
\end{answer}

\section{Equalizer and Coequalizer}

\begin{definition}{Minus-core of linear maps in vector spaces}
  Let $V, W$ be vector spaces over a field $K$. Let $s, t \from V \to W$ be linear functions.
  We define the linear function $s - t := \lambda v\in V. \; s(v) - t(v)$.
  Remember that for $g \from V \to W$ we have defined $core(g) := \set{ v \in V | g(v) = 0 }$.
  Now $core (s - t) = \set{v \in V | s(v) = t (v) }$.
\end{definition}

\begin{exercise}[Equalizer of vector spaces]
  Let
  $\xymatrix{
      X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y
  }$
  be a diagram in $Vect_K$ for a field $K$. Show that $core(s-t)$ is the equalizer for this diagram.
\end{exercise}

\begin{answer}
  \begin{proof}
    Let $i \from core(s-t) \to X$ be the inclusion function. Let
    $\xymatrix{
      A \ar[r]^f & X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y
    }$
    be a commuting diagram. Note that $f A \subseteq core(s-t)$.
    Thus we can define $\bar{f} \from A \to core(s-t)$ with $\bar{f}(a) := f(a)$.
    We see that $\forall a \in A: i(\bar{f}(a))=\bar{f}(a)=f(a)$, thus $i \of \bar{f} = f$.
    Let $\tilde{f} \from A \to core(s-t)$ be another map such that $i \of \tilde{f} = f$.
    Now $\tilde{f}(a) = i(\tilde{f}(a)) = i(\bar{f}(a)) = \bar{f}(a)$.
  \end{proof}
\end{answer}

\begin{exercise}
  Let $\parpair{X}{Y}{s}{t}$ be maps in some category.  Prove that $s = t$ if
  and only if the equalizer of $s$ and $t$ exists and is an isomorphism, if
  and only if the coequalizer of $s$ and $t$ exists and is an isomorphism.
\end{exercise}

\begin{answer}
  \begin{proof}
    We numerate the statements:
    \begin{itemize}
      \item  [I:]$s=t$
      \item [II:]The equalizer of $s$ and $t$ exists and is an isomorphism.
      \item[III:]The coequalizer of $s$ and $t$ exists and is an isomorphism.
    \end{itemize}
    We show that all statements are equivalent by showing $I \Leftrightarrow II$ and $I \Leftrightarrow III$.
    Note that $I \Leftrightarrow III$ is the dualized result of $I \Leftrightarrow II$, so we only need to show $I \Leftrightarrow II$.

    \begin{itemize}
      \item[``$\Rightarrow$'']
        Let $s=t$. $1_X$ is an isomorphism and we show that the fork
        $\xymatrix{
          X \ar[r]^{1_X} & X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y
        }$
        is the equalizer. Let
        $\xymatrix{
          A \ar[r]^f & X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y
        }$
        be a fork.
        Clearly $\bar{f} := f$ is the only map $\bar{f} \from A \to X$ such that the diagram
        \[ \xymatrix{
          A \ar[rd]^f \ar@{.>}[dd]|{\bar{f}} & & \\
          & X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \\
          X \ar[ru]_{1_X} & &
        } \]
        commutes.
      \item[``$\Leftarrow$'']
        Let
        $\xymatrix{
          E \ar[r]^i & X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y
        }$
        be the equalizer and $i$ a isomorphism. Then by commutativity we get:
        \begin{alignat*}{4}
          && s \of i &= t \of i \\
          \Rightarrow  &\qquad& s \of i \of i^{-1} &= t \of i \of i^{-1} \\
          \Rightarrow  &      & s &= t
        \end{alignat*}\qedhere
    \end{itemize}
  \end{proof}
\end{answer}

\begin{exercise}
  Let $X$ be a set and $f\from X \to X$ a map.  Describe the coequalizer of
  $\parpairi{X}{X}{f}{1}$ in $\Set$ as explicitly as possible.
\end{exercise}

\begin{answer}
  The definition of the coequalizer follows the scheme like in the talk.
  We define the relation $R := \set{ (1_X (x), f(x) \;\big|\; x \in X} = \set{ (x, f(x) \;\big|\; x \in X} = f$ seen as a functional relation.
  Now $\sim$ is the reflexive symmetric transitive closure of $R$.
  We define $i(x) := [x] := \set{ x' \in X | x \sim x'}$
  and the quotient $C := X / \sim \;=\; i(X)$.
  Now
  $ \xymatrix{
    X \ar@<.5ex>[r]^f \ar@<-.5ex>[r]_{1_X} & X \ar[r]^i & C
  } $
  is the coequalizer of $\parpairi{X}{X}{f}{1}$.

  We discover that $(x, y) \in \sim$ means that we can ``draw a line of forward and backward steps'' between $x$ and $y$.
  A ``forward step'' means that if we have $(x, y) \in \sim$ and $f(y)=z$ then $(x, z) \in \sim$.
  A ``backward step'' means that if we have $(x, y) \in \sim$ and $f(z)=y$ then $(x, z) \in \sim$.
  We also have $(x, x) \in \sim$ for every $x \in X$.
  More formally:
  With $\hat{R} := R \cup R^{-1}$ we have
  $\sim = \bigcup_{n\in\nat} {(\hat{R})^n}$
  where $(\hat{R})^n = \underbrace{\hat{R} \of \cdots \of \hat{R}}_{\text{$n$ times}}$ is the $n$th power relation over $\hat{R}$.
\end{answer}

\section{Pullbacks and Pushouts}

\begin{exercise}[Pushouts\,vs.\,coproducts]
  Show that if $Z$ is initial and the pushout square
  \[ \xymatrix{
    Z \ar[r]_s \ar[d]^t & X \ar[d]_{p_1} \\
    Y \ar[r]^{p_2} & P
  } \]
  is the pushout of
  $ \xymatrix{
    X & Z \ar[l]_s \ar[r]^t & Y \\
  } $
  then $P$ is the coproduct of $X$ and $Y$.

  Remark: Note that the dual statement holds for pullbacks of terminal objects and products.
\end{exercise}

\begin{answer}
  \begin{proof}
    We show that $S := P$ is the coproduct of $X$ and $Y$ together with the coprojection functions $p_1$ and $p_2$.
    Let
    $ \xymatrix{
      X \ar[r]^{f_1} & A & Y \ar[l]_{f_2}
    } $
    be a diagram in $\cat{A}$.
    We have to show that there is a unique map $\bar{f}' \from P \to A$ such that
    \[ \xymatrix{
      X \ar[rdd]_{f_1}\ar[rd]^{p_1} & & Y \ar[ldd]^{f_2} \ar[ld]_{p_2} \\
      & P \ar@{.>}[d]|{\bar{f}'} & \\
      & A &
    } \]
    commutes. We know that the diagram
    \[ \xymatrix{
      Z \ar[r]_s \ar[d]^t & X \ar[d]_{f_1} \\
      Y \ar[r]^{f_2} & A
    } \]
    commutes because $Z$ is initial.
    Now because this square is then a pushout square there exists a unique map $\bar{f}\from P \to A$ such that
    \[ \xymatrix{
      Z \ar[r]_s \ar[d]^t & X \ar[d]_{p_1} \ar[ddr]^{f_1} & \\
      Y \ar[r]^{p_2} \ar[drr]_{f_2}& P \ar@{.>}[dr]|{\bar{f}} & \\
      & & A  \\
    } \]
    commutes. We see that we can choose $\bar{f}' := \bar{f}$.
    Suppose we have another $\tilde{f} \from P \to A$ such that
    \[ \xymatrix{
      X \ar[rdd]_{f_1}\ar[rd]^{p_1} & & Y \ar[ldd]^{f_2} \ar[ld]_{p_2} \\
      & P \ar@{.>}[d]|{\tilde{f}} & \\
      & A &
    } \]
    commutes. Then
    \[ \xymatrix{
      Z \ar[r]_s \ar[d]^t & X \ar[d]_{p_1} \ar[ddr]^{f_1} & \\
      Y \ar[r]^{p_2} \ar[drr]_{f_2}& P \ar@{.>}[dr]|{\tilde{f}} & \\
      & & A  \\
    } \]
    would also commute (because $Z$ is initial) but $\bar{f}$ is unique, hence $\bar{f} = \tilde{f}$.
  \end{proof}

\end{answer}

\begin{exercise}
  Take objects and maps $\xymatrix@1{E \ar[r]^i &X \ar@<.5ex>[r]^{f}
  \ar@<-.5ex>[r]_g &Y}$ in some category.  If this is an equalizer, is the
  square
  \[ \xymatrix{
    E \ar[r]^i \ar[d]_i &
    X \ar[d]^g \\
    X \ar[r]_f &
    Y
  } \]
  necessarily a pullback?
  What about the converse?  Give proofs or counterexamples.
\end{exercise}

\begin{answer}
  \begin{itemize}
    \item[``$\Rightarrow$'']
      \emph{Counterexample:}
      Let $\cat{A} := \Set, E := \emptyset, X := \nat, Y := \mathbb{B}$.
      Define $f(n) := \text{even}(n)$ and $g(n) := \text{odd}(n)$. $i$ is the (empty) inclusion function.
      Then
      $\xymatrix@1{E \ar[r]^i &X \ar@<.5ex>[r]^{f}\ar@<-.5ex>[r]_g &Y}$
      is clearly the equalizer because no number is even and odd at the same time.
      Now let $A := \nat$, $f_1(n) := n$ and $f_2(n) := n+1$.
      Then the diagram
      \[ \xymatrix{
        A \ar[r]_{f_2} \ar[d]^{f_1} & Y \ar[d]_{g} \\
        X \ar[r]^f & Z
      } \]
      commutes because $\text{even}(n) \Leftrightarrow \text{odd}(n+1)$.
      But there exists no function $\bar{f} \from \nat \to \emptyset = E$ at all.
    \item[``$\Leftarrow$'']
      \begin{proof}
        Let
        \[ \xymatrix{
          E \ar[r]^i \ar[d]_i &
          X \ar[d]^g \\
          X \ar[r]_f &
          Y
        } \]
        be the pullback of the diagram
        $ \xymatrix{
          X \ar[r]^f & Y & X\ar[l]_g \\
        } $.
        We show that $\xymatrix@1{E \ar[r]^i &X \ar@<.5ex>[r]^{f} \ar@<-.5ex>[r]_g &Y}$ is the equalizer of $\xymatrix@1{X \ar@<.5ex>[r]^{f} \ar@<-.5ex>[r]_g &Y}$.
        First we note that the diagram $\xymatrix@1{E \ar[r]^i &X \ar@<.5ex>[r]^{f} \ar@<-.5ex>[r]_g &Y}$ is indeed a fork as it commutes by assumption.
        Let
        $ \xymatrix{
          A \ar[r]^f & X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y
        } $
        be a fork. Then
        \[ \xymatrix{
          A \ar[r]_{f} \ar[d]^{f} & X \ar[d]_{t} \\
          X \ar[r]^s & Y
        } \]
        commutes. Therefore there exists a unique map $\bar{f}\from A \to P$ such that
        \[ \xymatrix{
          A \ar[drr]^{f} \ar[ddr]_{f} \ar@{.>}[dr]|{\bar{f}} & & \\
          & P \ar[r]_{i} \ar[d]^{i} & X \ar[d]_{t} \\
          & X \ar[r]^s & Y
        } \]
        commutes. We show that this $\bar{f}$ is the only map such that
        \[ \xymatrix{
          A \ar[rd]^f \ar@{.>}[dd]|{\bar{f}} & & \\
          & X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \\
          P \ar[ru]_i & &
        } \]
        commutes. First it is clear that this diagram commutes for $\bar{f}$. Now let $\tilde{f}$ be another map such that
        \[ \xymatrix{
          A \ar[rd]^f \ar@{.>}[dd]|{\tilde{f}} & & \\
          & X \ar@<.5ex>[r]^s \ar@<-.5ex>[r]_t & Y \\
          P \ar[ru]_i & &
        } \]
        commutes. But then
        \[ \xymatrix{
          A \ar[drr]^{f} \ar[ddr]_{f} \ar@{.>}[dr]|{\tilde{f}} & & \\
          & P \ar[r]_{i} \ar[d]^{i} & X \ar[d]_{t} \\
          & X \ar[r]^s & Y
        } \]
        also commutes. By the assumption that $\bar{f}$ is unique we have $\bar{f} = \tilde{f}$. \qedhere
      \end{proof}
  \end{itemize}
\end{answer}

\begin{exercise}
  Take a commutative diagram
  \[ \xymatrix{
    \cdot \ar[r] \ar[d] & \cdot \ar[r] \ar[d] & \cdot \ar[d] \\
    \cdot \ar[r] & \cdot \ar[r] & \cdot
  } \]
  in some category.  Suppose that the right-hand square is a pullback.  Show
  that the left-hand square is a pullback if and only if the outer rectangle is
  a pullback.
\end{exercise}

\begin{answer}
  \begin{proof}
    We give names to the arrows and objects:
    \[ \xymatrix{
      A \ar[r]^h \ar[d]^j & B \ar[r]^i \ar[d]^k & C \ar[d]^l \\
      D \ar[r]^m & E \ar[r]^n & F
    } \]
    Now let the pullback square
    \[ \xymatrix{
      B \ar[r]^i \ar[d]^k & C \ar[d]^l \\
      E \ar[r]^n & F
    } \]
    be the pullback of $\xymatrix{E \ar[r]^n & F & C \ar[l]_l}$.
    We show both directions of the equivalence:
    \begin{itemize}
      \item[``$\Rightarrow$'']
        Let the pullback square
        \[ \xymatrix{
          A \ar[r]^{i \of h} \ar[d]^j & C \ar[d]^l \\
          D \ar[r]^{n \of m} & F
        } \]
        be the pullback of $\xymatrix{D \ar[r]^{n \of m} & F & C \ar[l]_l}$.
        Let
        \[ \xymatrix{
          Z \ar[r]^{f_1} \ar[d]^{f_2} & B \ar[d]^k \\
          D \ar[r]^{m} & E
        } \]
        be another pullback square for the left-hand square. Now
        \[ \xymatrix{
          Z \ar[r]^{f_1} \ar[d]^{f_2} & B \ar[r]^{i} \ar[d]^k & C \ar[d]^l \\
          D \ar[r]^{m} & E \ar[r]^n & F
        } \]
        commutes. Now because the outer rectangle is a pullback there is a unique map $\bar{f} \from Z \to A$, such that
        \[ \xymatrix{
          Z \ar@/^/[drrr]^{i \of f_1}
            \ar@{.>}[dr]|{\bar{f}}
            \ar@/_/[ddr]_{f_2} & & \\
            & A \ar[r]_h \ar[d]^j & B \ar[r]_{i} & C \ar[d]_l \\
            & D \ar[r]^{m} & E \ar[r]^n & F
        } \]
        commutes. We show that this $\bar{f}$ is the unique map such that
        \[ \xymatrix{
          Z \ar@/^/[drr]^{f_1}
            \ar@{.>}[dr]|{\bar{f}}
            \ar@/_/[ddr]_{f_2} \\
          & A \ar[r]^h \ar[d]^j & B \ar[d]^k \\
          & D \ar[r]^m & E
        } \]
        commutes. We have to show $j \of \bar{f} = f_2$ and $h \of \bar{f} = f_1$. The first equatation holds by assumption.

        Now we define two maps
        \begin{alignat*}{4}
          g_1 &\from Z \to C&  \qquad g_1 &:= i \of f_1 \\
          g_2 &\from Z \to E&         g_2 &:= m \of j \of \bar{f}
        \end{alignat*}
        and show that the diagram
        \[ \xymatrix{
          Z \ar[r]^{g_1} \ar[d]^{g_2} & C \ar[d]^l \\
          E \ar[r]^n & F
        } \]
        commutes and is therefore a pullback square of the right-hand square:
        \begin{alignat*}{2}
          l \of g_1
          &= l \of i \of f_1 \\
          &= l \of i \of h \of \bar{f} \\
          &= n \of m \of j \of \bar{f} \\
          &= n \of g_2
        \end{alignat*}
        Now because the right-hand square is a pullback there is a unique map $\bar{g} \from Z \to B$, such that
        \[ \xymatrix{
          Z \ar@/^/[drr]^{g_1}
            \ar@{.>}[dr]|{\bar{g}}
            \ar@/_/[ddr]_{g_2} \\
          & B \ar[r]^i \ar[d]^k & C \ar[d]^l \\
          & E \ar[r]^n & F
        } \]
        commutes.
        By showing that this diagram comutes for $\bar{g} = f_1$ and for $\bar{g} = h \of \bar{f}$ we show $h \of \bar{f} = f_2$.
        \begin{itemize}
          \item[1.]$\bar{g} = f_1$: We have $i \of f_1 = g_1$ by definition and $k \of f_1 = m \of f_2 = m \of j \of \bar{f} = g_2$.
          \item[2.]$\bar{g} = h \of \bar{f}$: We have $(i \of h) \of \bar{f} = i \of f_2$ and $(k \of h) \of \bar{f} = (m \of j) \of \bar{f}$.
        \end{itemize}

        We have shown that the diagram
        \[ \xymatrix{
          Z \ar@/^/[drr]^{f_1}
            \ar@{.>}[dr]|{\bar{f}}
            \ar@/_/[ddr]_{f_2} \\
          & A \ar[r]^h \ar[d]^j & B \ar[d]^k \\
          & D \ar[r]^m & E
        } \]
        commutes for the given $\bar{f}$. We now have to prove that this $\bar{f}$ is unique.
        We assume a map $\tilde{f} \from Z \to A$ such that
        \[ \xymatrix{
          Z \ar@/^/[drr]^{f_1}
            \ar@{.>}[dr]|{\tilde{f}}
            \ar@/_/[ddr]_{f_2} \\
          & A \ar[r]^h \ar[d]^j & B \ar[d]^k \\
          & D \ar[r]^m & E
        } \]
        commutes. But than we have another map such that the diagram for the outer rectangle commutes:
        \[ \xymatrix{
          Z \ar@/^/[drrr]^{i \of f_1}
            \ar@{.>}[dr]|{\tilde{f}}
            \ar@/_/[ddr]_{f_2} & & \\
            & A \ar[r]_h \ar[d]^j & B \ar[r]_{i} & C \ar[d]_l \\
            & D \ar[r]^{m} & E \ar[r]^n & F
        } \]
        We conclude that $\bar{f} = \tilde{f}$.


      \item[``$\Leftarrow$'']
        Let the pullback square
        \[ \xymatrix{
          A \ar[r]^h \ar[d]^j & B \ar[d]^k \\
          D \ar[r]^m & E
        } \]
        be the pullback of $\xymatrix{D \ar[r]^m & E & B \ar[l]_k}$.
        Let
        $ \xymatrix{
          D & Z \ar[l]_{f_2} \ar[r]^{f_1} & C
        } $
        be a diagram such that
        \[ \xymatrix{
          Z \ar[r]^{f_1} \ar[d]_{f_2} & C \ar[d]^l \\
          D \ar[r]^{n\of m} & F
        } \]
        commutes. We note that
        \[ \xymatrix{
          Z \ar[r]^{f_1} \ar[d]_{m \of f_2} & C \ar[d]^l \\
          E \ar[r]_n & F
        } \]
        is thus a pullback square of the right-hand square.
        Thus we have a unique map $\bar{f} \from Z \to B$ such that
        \[ \xymatrix{
          Z \ar@/^/[drr]^{f_1}
            \ar@{.>}[dr]|{\bar{f}}
            \ar@/_/[ddr]_{m \of f_2} \\
          & B \ar[r]^i \ar[d]^k & C \ar[d]^l \\
          & E \ar[r]^n & F
        } \]
        commutes. Then
        \[ \xymatrix{
          Z \ar[r]^{\bar{f}} \ar[d]^{f_2} & B \ar[d]^k \\
          D \ar[r]^m && E
        } \]
        is a pullback square of the left-hand square.
        Thus we have a unique map $\bar{g} \from Z \to A$ such that
        \[ \xymatrix{
          Z \ar@/^/[drr]^{\bar{f}}
            \ar@{.>}[dr]|{\bar{g}}
            \ar@/_/[ddr]_{f_2} \\
          & A \ar[r]^h \ar[d]^j & B \ar[d]^k \\
          & D \ar[r]^m & E
        } \]
        commutes. We show that this $\bar{g}$ is the unique map such that
        \[ \xymatrix{
          Z \ar@/^/[drr]^{f_1}
            \ar@{.>}[dr]|{\bar{g}}
            \ar@/_/[ddr]_{f_2} \\
          & A \ar[r]^{i \of h} \ar[d]^j & C \ar[d]^l \\
          & D \ar[r]^{n \of m} & F
        } \]
        commutes.
        The diagram comutes for $\bar{g}$:
        \begin{align*}
          i \of h \of \bar{g} = i \of \bar{f} = f_1 \\
          j \of \bar{g} = f_2
        \end{align*}
        Let $\tilde{g}$ be another map such that
        \[ \xymatrix{
          Z \ar@/^/[drr]^{f_1}
            \ar@{.>}[dr]|{\tilde{g}}
            \ar@/_/[ddr]_{f_2} \\
          & A \ar[r]^{i \of h} \ar[d]^j & C \ar[d]^l \\
          & D \ar[r]^{n \of m} & F
        } \]
        commutes. To show $\bar{g} = \tilde{g}$ it is because of the uniqueness enough to show that
        \[ \xymatrix{
          Z \ar@/^/[drr]^{\bar{f}}
            \ar@{.>}[dr]|{\tilde{g}}
            \ar@/_/[ddr]_{f_2} \\
          & A \ar[r]^h \ar[d]^j & B \ar[d]^k \\
          & D \ar[r]^m & E
        } \]
        commutes.
        We already have $j \of \tilde{g} = f_2$ and only have to show $h \of \tilde{g} = \bar{f}$.
        Because of $\bar{f} = h \of \bar{g}$ it suffices to show $h \of \tilde{g} = h \of \bar{g}$.
        The commuting square
        \[ \xymatrix{
          Z \ar[r]^{i \of h \of \tilde{g}} \ar[d]_{m \of j \of \tilde{g}} & C \ar[d]^l \\
          E \ar[r]^n & F
        } \]
        is another pullback square for the right-hand square.
        Thus we have a unique map $\bar{x} \from Z \to B$ such that
        \[ \xymatrix{
          Z \ar@/^/[drr]^{i \of h \of \tilde{g}}
            \ar@{.>}[dr]|{\bar{x}}
            \ar@/_/[ddr]_{m \of j \of \tilde{g}} \\
          & B \ar[r]^i \ar[d]^k & C \ar[d]^l \\
          & E \ar[r]^n & F
        } \]
        commutes.
        We get $h \of \tilde{g} = h \of \bar{g}$ when we show that this diagram commutes for $\bar{x} = h \of \tilde{g}$ and $\bar{x} = h \of \bar{g}$.

        \begin{itemize}
          \item[1.]$\bar{x} = h \of \tilde{g}$:
            We have $i \of h \of \tilde{g} = i \of h \of \tilde{g}$ and $m \of j \of \tilde{g} = k \of h \of \tilde{g}$.
          \item[2.]$\bar{x} = h \of \bar{g}$:
            We have $i \of h \of \bar{g} = i \of f_1 = i \of h \of \tilde{g}$. Furthermore we have
            \begin{alignat*}{3}
              & &\quad&  m \of j \of \tilde{g} \\
              &=&& m \of f_2 \of \tilde{g} \\
              &=&& m \of j \of \bar{g} \\
              &=&& k \of h \of \bar{g}.
            \end{alignat*}
        \end{itemize}
        Thus $h \of \tilde{g} = h \of \bar{g}$. Hence the outer rectangle is a pullback.\qedhere
    \end{itemize}
  \end{proof}
\end{answer}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
% vim: ts=2 sts=2 sw=2 expandtab
